---
title: "R Basic Concepts: <br>Sophisticated subsetting"
company: "`r if(exists('info')) info$company`"
author: "`r if(exists('info')) info$author`"
date: "`r if(exists('info')) info$date`"
output:
  ioslides_presentation: 
    css: styles.css
    logo: img/ZEIF.png
    widescreen: yes
    df_print: kable
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
study <- data.frame(
  sen    = c(0, 1, 0, 1, 0, 1),
  gender = c("M", "M", "F", "M", "F", "F"),
  age    = c(12, 13, 11, 10, 11, 14),
  IQ     = c(90, 85, 90, 87, 99, 89)
)
```

# Sophisticated selections

## Relational operators {.build}

Relational operators compare two values and return a logical value (`TRUE` or `FALSE`)

|Operator|Relation|Example|
|----|----------------|--------|
|`==`|is identical|x == y|
|`!=`|is not identical|x != y|
|`>`|is greater|x > y|
|`>=`|is greater or identical|x >= y|  
|`<`|is less|x < y|
|`<=`|is less or identical|x <=  y|  

## Examples {.build}

```{r}
7 > 2
7 <=  10
5 == 4
5 != 6
```

## Relational vectors and characters

Only `==` and `!=` can be applied to non numerical objects:

```{r}
"Hamster" == "Mouse"
"Hamster" != "Mouse"
```

## Relational operators and vectors {.smaller}

```{r}
age <- c(12, 4, 3, 8, 4, 2, 1)
age < 5
```

```{r echo = FALSE}
data.frame(age = age, 'age < 5' = age < 5, check.names = FALSE)
```

## Using logical vectors to select values

When you put a logical vector within square brackets `[ ]` after an object, all elements of that object with a `TRUE` in the logical vector are selected:

```{r}
age <- c(12, 4, 3, 8)
x <- age > 5
x
age[x]
```

## Using logical vectors to select values

```{r eval=FALSE}
age <- c(12, 4, 3, 8)
x <- age > 5
age[x]
```

```{r echo = FALSE}
data.frame(age = age, x = x, 'Select?' = ifelse(x, "**select**", "drop"), "Result" = ifelse(x, age, ""), check.names = FALSE)
```

## Task {.build}

Create a new vector `friends <- c(4, 5, 6, 3, 7, 2, 3)`. Show all values of that vector `>= 4`.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution 

Create a new vector `friends <- c(4, 5, 6, 3, 7, 2, 3)`. Show all values of that vector `>= 4`.

```{r}
friends <- c(4, 5, 6, 3, 7, 2, 3)
friends[friends >= 4]
```


## which() {.build .smaller}

The `which()` functions gives the **indices** of the elements that are `TRUE`.  
It takes a *logical vector* as an *argument*.

```{r}
x <- c(TRUE, FALSE, FALSE, TRUE)
which(x)

age <- c(12, 4, 3, 8)
x <- age < 5
x
which(x)
```

---

```{r eval = FALSE}
age <- c(12, 4, 3, 8)
x <- age < 5
x
which(x)
```

```{r echo = FALSE}
data.frame('Index' = 1:length(x), '`age`' = age, '`x <- age < 5`' = x, '`which(x)`' = ifelse(x, 1:length(x), ""), check.names = FALSE)
```

## Task {.build}

Create a vector `x <- c(1, 4, 5, 3, 4, 5)` and identify:  
1. Which elements are larger or equal than three?     
2. Create a new vector from `x` containing all elements that are **not** four. Note: Use the `which()` function for this task.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Create a vector `x <- c(1, 4, 5, 3, 4, 5)` and identify:  
1. Which elements are larger or equal than three?     
2. Create a new vector from `x` containing all elements that are **not** four. Note: Use the `which()` function for this task.


```{r}
x <- c(1, 4, 5, 3, 4, 5)
which(x >= 3)
y <- x[which(x != 4)]
y
```

## Selecting cases with logical vectors {.build}

Logical vectors can also be appplied to *data frames* for selecting cases:

```{r}
# Either directly:
study_no_sen <- study[study[["sen"]] == 0, ]
study_no_sen

# Or using the which() function
filter <- which(study[["sen"]] == 0)
study_no_sen <- study[filter, ]
```

## Task {.build}

Calculate the `mean` of `IQ` for students with and without sen.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Calculate the `mean` of `IQ` for students with and without sen.

```{r}
filter <- which(study[["sen"]] == 0)
mean(study[["IQ"]][filter])

filter <- which(study[["sen"]] == 1)
mean(study[["IQ"]][filter])
```

## Logical Operations

Logical operations are applied to logical values. 

|Operator| Operation|Example|Results|
|--------|----------|-------|-------|
|`!`|Not|`! x`|`TRUE when x = FALSE and FALSE when x = TRUE`|
|`&`|AND|`x & y`|`TRUE when x and y are TRUE else FALSE`|
|`|`|OR|`x | y`|`TRUE when x or y is TRUE else FALSE`|

<br>
<div class="emphasized">
Note: To get the | sign:  
On a german Mac keyboard press: option + 7  
On a german Windows keyboard press: AltGr + <  
</div>

## Example {.build .smaller}

```{r}
x <- TRUE
y <- FALSE
```

```{r}
!x
!y
x & y
x | y
```

## Logical Operator with vectors {.build .smaller}

When applied to **two** vectors, logical operations result in a new vector.  
Operations are applied to each element one by one.

```{r}
x <- c(TRUE, FALSE, TRUE,  FALSE)
y <- c(TRUE, FALSE, FALSE, TRUE)
```

```{r}
!x
x & y
x | y
```

## Task {.build}

Create two vectors:

`glasses <- c(TRUE, TRUE, FALSE, TRUE, FALSE)`  
`hyperintelligent <- c(TRUE, FALSE, FALSE, TRUE, FALSE)`

Determine for each element whether 'glasses' and 'hyperintelligent' are TRUE at the same time.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 


## Task - solutions

Create two vectors:

`glasses <- c(TRUE, TRUE, FALSE, TRUE, FALSE)` 
`hyperintelligent <- c(TRUE, FALSE, FALSE, TRUE, FALSE)`

Determine for each element whether 'glasses' and 'hyperintelligent' are TRUE at the same time.

```{r}
glasses <- c(TRUE, TRUE, FALSE, TRUE, FALSE)
hyperintelligent <- c(TRUE, FALSE, FALSE, TRUE, FALSE)
glasses & hyperintelligent
```

---

```{r echo = FALSE}
data.frame('glasses' = glasses, 'hyperintelligent' = hyperintelligent,  'glasses & hyperintelligent' = glasses & hyperintelligent, check.names = FALSE)
```

## Combining logical and relational operators {.build}

```{r}
age <- c(12, 4, 3, 8, 4, 2, 1, 7, 4)
gender <- c(0, 1, 0, 1, 0, 0, 0, 0, 1)
age > 4
gender == 0
age > 4 & gender == 0
```

## Task {.build .smaller}

Create a vector  
`income <- c(5000, 4000, 3000, 2000, 1000)` and a vector  
`happiness <- c(20, 35, 30, 10, 50)`.  
Use *relational* and *logical* operations to determine for each element whether the `income` is larger than `2500` and at the same time `happiness` is above `25`.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

```{r}
income <- c(5000, 4000, 3000, 2000, 1000)
happiness <- c(20, 35, 30, 10, 50)
income > 2500 & happiness > 25
```

## Task - solution {.smaller}

Create a vector  
`income <- c(5000, 4000, 3000, 2000, 1000)` and a vector  
`happiness <- c(20, 35, 30, 10, 50)`.  
Use *relational* and *logical* operations to determine for each element whether the `income` is larger than `2500` and at the same time `happiness` is above `25`.

---

```{r eval = FALSE}
income <- c(5000, 4000, 3000, 2000, 1000)
happiness <- c(20, 35, 30, 10, 50)
income > 2500 & happiness > 25
```

```{r echo = FALSE}
kable(data.frame('income' = income, 'happiness' = happiness, 'income > 2500' = income > 2500, 'happiness > 25' = happiness > 25, 'income > 2500 &<br>happiness > 25' = income > 2500 & happiness > 25, check.names = FALSE))

```

## Subsetting data frames with logical and relational operators {.build .smaller}

```{r}
study
```

---

```{r}
filter <- study[["sen"]] == 1 & study[["gender"]] == "M"
study[filter, ]
```

## Task

Use the `ChickWeight` data frame for the following task.  
The data set is already included in R.

1. Look into the data set with `?ChickWeight`.
2. Get all variable names of the data frame with the `names()` function (`names(ChickWeight)`).
3. Select cases from **ChickWeight** with `Diet == 1` and `Time < 16`.
4. For these cases, calculate the correlation between `weight` and `Time`. Note: Use the `cor()` function (e.g., `cor(x, y)`)
5. Repeat steps 3 and 4 for `Diet == 4`.
6. What can you see?

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

---

```{r}
filter <- ChickWeight[["Diet"]] ==  1 & ChickWeight[["Time"]] < 16
diet1 <- ChickWeight[filter,]
cor(diet1[["weight"]], diet1[["Time"]])
filter <- ChickWeight[["Diet"]] ==  4 & ChickWeight[["Time"]] < 16
diet4 <- ChickWeight[filter,]
cor(diet4[["weight"]], diet4[["Time"]])
```

