---
title: "The Rmarkdown silly example"
author: "Jürgen Wilbert"
date: "12.12.2019"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

# header one
## header two
### header three

This is *italic* this is **bold**, and ***this is both!!!***.

An unordered list:

- Helmut
- Gerhard
- Angela

An ordered list:

1. Bush
2. Obama
3. Trump

Combined:
  
1. Sour
    - Lime
    - Envy
2. Sweet
    - Honey
    - love
  

# Example

Now follow some examples. We start with some analyses on Star Wars characters.

The package `dplyr` has a database included.

## Starwars

```{r include=FALSE}
library(dplyr) # needed for the starwars database
library(knitr) # needed for the kable() function
```

The Starwars data set contains the following variables:

```{r echo = FALSE}
names(starwars)
```

Do different species have different hights?

```{r echo = FALSE, fig.cap = "Hight distribution of Star Wars species"}
dat <- starwars[starwars$species %in% c("Droid", "Human", "Wookiee"),]
boxplot(dat$height ~ dat$species)
```

Is there a gender bias in Star Wars?

```{r echo = FALSE}
tab <- table(starwars$gender)
kable(tab, caption = "Gender distribution of Star Wars characters")
```

And finally, from which homeworlds do the characters come from?

```{r echo = FALSE}
filter <- names(which(table(starwars$species) > 2))
dat <- starwars[starwars$species %in% filter,]
tab <- table(dat$homeworld, dat$species)
tab <- rbind(tab, total = margin.table(tab, 2))
kable(tab)
```


