---
title: "Basic statistics"
format: revealjs
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
```

## Goals  {background-image="img/bg-goal.jpg" background-opacity=0.3}

-   You know some functions for basic statistical analyses.
-   You can create contingency tables.
-   You learn to integrate filtering, selecting, grouping, and subsetting of data frames with functions for statistical analyses.

# Descriptive statistics {background-image="img/bg-h1-3.jpg"}

## Some basic statistical functions

-   `min()`, `max()`: Minimum and maximum
-   `mean()`, `median()`: Mean and median
-   `sd()`, `var()`: Standard deviation and variance
-   `mad()`: Median absolute deviation
-   `quantile()`: Percentile / quantile

## Task

-   Take the `mtcars` dataset
-   Create a new variable `lpk` (liters per 100km) from `mpg`\
    (Formula: `lpk = 282.5 / mpg`)
-   Calculate grouped by cylinders (`cyl`): mean, sd, median, mad, min, max of `lpk`
-   Round all values by one decimal
-   Hint: Use the `tidyverse` or `dplyr` library

{{< include _stop.qmd >}}

##  {.smaller}

```{r}
mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(cyl) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  round(1) 

```

## 

What is happening here?

```{r}
mtcars <- mutate(mtcars, 
  transmission = factor(am, labels = c("Manual", "Automatic"))
)
```

## 

Now, lets calculate the stats for the transmission type:

```{r}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) 

```

## 

... and round the results

```{r eval = FALSE}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  round(1)

```

```{r echo = FALSE}
cat("Error in Math.data.frame(list(transmission = 1:2, mean = c(17.3586131968484, : 
non-numeric-alike variable(s) in data frame: transmission")
```

ups :-( ... what went wrong?

## 

Solution: only `round` variables that are numeric:

-   `is.numeric()` returns a TRUE or FALSE for a vector (`is.numeric(1:5)`; `is.numeric(c("A", "B"))`)\
-   `across()` specifies for `mutate()`which variable to select.

`mutate(across(.cols, .fns, ... ))`:

-   .cols := columns so select\
-   .fns := a function to apply\
-   ... := arguments to that function

## 

```{r}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  mutate(across(where(is.numeric), round, 1))

```

# Tables {background-image="img/bg-h1-3.jpg"}

## Tables in base R

-   `table()`: Shows frequencies of nominal scaled variables.
-   `prop.table()`: Calculates proportions from frequency tables.
-   `addmargins()`: Adds margins to tables.

## Example one dimensional {.smaller}

```{r}
tab <- table(mtcars$cyl) # frequencies
tab
prop.table(tab) # proportions
prop.table(tab) * 100 # percentages 
```

## Example two dimensional {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
tab
prop.table(tab) * 100
```

## Example two dimensional with percentages by rows and columns {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
prop.table(tab, margin = 1) * 100  # sum of each row is 100%
prop.table(tab, margin = 2) * 100 # sum of each column is 100%
```

## Example with added margins {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
tab <- prop.table(tab, margin = 2) * 100
addmargins(tab)
```

## Task

-   Take the `heights` dataset from the `dslabs` library.\
-   Create a cross table depicting: The percentage of females and males that are above and below a height of 170 cm.

## 

```{r}
library(dslabs)

heights$category[heights$height * 2.54 <= 170] <- "% <= 170 cm"
heights$category[heights$height * 2.54 > 170] <- "% > 170 cm"

tab <- table(heights$sex, heights$category)
tab <- prop.table(tab, margin = 1) * 100
tab <- round(tab, 1)
tab
```

## Tables with dplyr

```{r}
mtcars %>% 
  group_by(cyl, am) %>% 
  summarise(n = n())

```

## Shortcut for simple count tables

```{r}
# Because it is shorter, I don't use pipes here (but I could!)
count(mtcars, cyl) 
count(mtcars, cyl, am) 
```

## Tables with dplyr: `pivot_wider()` {.smaller}

`pivot_wider()`:\
creates separate variables from the levels of a factor variable and the values of a second variable

##  {.smaller}

```{r}
mtcars %>% count(cyl, am)
```

```{r}
mtcars %>% count(cyl, am) %>%
  pivot_wider(names_from = "am", values_from = "n")

```

## Tables with dplyr: A bit nicer!

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  count(cyl, am) %>% 
  pivot_wider(names_from = "am", values_from = "n") %>%
  rename("Cylinders" = "cyl")

```

## Task

Recreate the previous task. This time using dplyr:

-   Take the `heights` dataset from the `dslabs` library.\
-   Create a cross table depicting: The percentage of females and males that are above and below a height of 170 cm.

This is a hard task ...

Hint 1: `ifelse(height * 2.54 <= 170, "size_a", "size_b")`\
Hint 2:\
`size_a / (size_a + size_b) * 100`\
`size_b / (size_a + size_b) * 100`\
are the row proportions.

## 

```{r}
library(dslabs)

heights %>%
  mutate(category = ifelse(height * 2.54 <= 170, "size_a", "size_b")) %>% 
  count(sex, category) %>%
  pivot_wider(names_from = "category", values_from = "n") %>%
  mutate(
    "% <= 170" = size_a / (size_a + size_b) * 100,
    "% > 170" = size_b / (size_a + size_b) * 100,
    across(where(is.numeric), round, 1)
  ) %>%
  select(1, 4, 5)

```

## Tables with dplyr: For values of a third variable

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(
    n = n(), 
    M = mean(mpg), 
    SD = sd(mpg)
  ) 
```

## 

`pivot_wider()` can take values from several variables:

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(n = n(), M = mean(mpg), SD = sd(mpg)) %>%
  pivot_wider(names_from = "am", values_from = c("n", "M", "SD")) %>%
  round(1)

```

## 

Set argument `names_vary = "slowest"` for a different ordering of variables:

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(n = n(), M = mean(mpg), SD = sd(mpg)) %>%
  pivot_wider(names_from = "am", values_from = c("n", "M", "SD"), names_vary = "slowest") %>%
  round(1)

```

## Task

-   Take the `storms` dataset and calculate n and mean of the wind speed (`wind`) by month (`month`) and storm classification (`status`) in a crosstable.

## 

```{r}
storms %>% 
  group_by(month, status) %>%
  summarise(
    n = n(),
    M = mean(wind, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = "status", values_from = c("n", "M")) %>%
  round()
  
```

## Design example:

Replicate and try to understand the following table example:

```{r eval = FALSE}
library(kableExtra)
tab <- mtcars %>% 
  group_by(cyl, am) %>% 
  summarise(n = n(), M = mean(mpg), SD = sd(mpg)) %>%
  pivot_wider(names_from = "am", values_from = c("n", "M", "SD"), names_vary = "slowest") 

names(tab) <- c("Cylinders", "n", "M", "SD", "n", "M", "SD")

kable(tab, 
  caption =  "Table 1.<br>Miles per gallon by cylinders and gearshift", 
  digits = 1, 
  full_width = FALSE
) %>% 
  kable_classic() %>%
  add_header_above(c(" " = 1, "Automatic" = 3, "Manual" = 3))
```

## 

```{r echo = FALSE}
library(kableExtra)
tab <- mtcars %>% 
  group_by(cyl, am) %>% 
  summarise(n = n(), M = mean(mpg), SD = sd(mpg)) %>%
  pivot_wider(names_from = "am", values_from = c("n", "M", "SD"), names_vary = "slowest") 
names(tab) <- c("Cylinders", "n", "M", "SD", "n", "M", "SD")
kable(tab, caption =  "Table 1.<br>Miles per gallon by cylinders and gearshift", digits = 1) %>% kable_classic() %>%
  add_header_above(c(" " = 1, "Automatic" = 3, "Manual" = 3))
```

# Inferential statistics {background-image="img/bg-h1-3.jpg"}

## Some statistical functions:

-   `t.test()`: Calculating a t-test.

-   `wilcox.test()`: Calculating Wilcox test / U-Test.

-   `chisq.test()`: Calculating a Pearson $X^2$-test.

-   Some more we will not address today:

    -   `lm()`: Regression analyses
    -   `cor.test()`: Calculating a correlation test.
    -   `binom.test()`: Binomial test.
    -   `fisher.test()`: Fisher exact test for count data.
    -   `ks.test()`: One and two sample Kolmogorov-Smirnov Tests.
    -   `shapiro.test()`: Shapiro-Wilk Normality Test.
    -   `aov()`: Analysis of variance

## `chisq.test()`

-   *Pearson's* $X^2$ test is a very versatile test to calculate whether a distribution of observed frequencies equates expected frequencies.\
-   A very common application is to test whether frequencies of observations in two variables are related. (E.g. number of patients that improved in a treatment vs. control group).
-   The `chisq.test()` functions takes a two dimensional frequency table and calculates a $X^2$ test.

## Example with data from an intervention study {.smaller}

```{r}
# get an external data example
dat <- read.csv("https://goo.gl/j6lRXD")
# Create a two distribution table
tab <- table(dat$treatment, dat$improvement)
tab
# Test for non random distribution:
chisq.test(tab) # Alternatively: chisq.test(dat$treatment, dat$improvement)
```

Significantly more patients improved in the treatment condition ($X^2(1)=4.7, p < .05$)

## Task

-   Take the `starwars` dataset from the `tidyverse` package.
-   Use `table()` to get the distribution `eye_color` by `hair_color`
-   Apply the `chisq.test()` to calculate a X² test to test for an even distribution.

{{< include _stop.qmd >}}

## 

```{r warning=FALSE}
tab <- table(starwars$eye_color, starwars$hair_color)
chisq.test(tab)
```

## `t.test()`

-   *Student's t-test* analysis whether two samples originate from the same normal distribution.\

-   It is used to test for mean differences in two groups.\

-   ***Arguments*** of the `t.test()` function:

    -   `x` and `y`: Each variable provides data from a samples.
    -   `formula`: If you have one vector with all data (e.g. `values`) and a second vector with grouping information (e.g. `group`) use: `values ~ group`.
    -   `data`: If you work with `formula`: The name of a data frame.
    -   `paired`: If set `TRUE`, the test assumes repeated measures of one sample instead of two independent samples.

## Example

Does one of two drugs increase hours of sleep better?

`extra`: change in sleep duration, `group` : drug given

```{r}
sleep
```

## Example

```{r eval = FALSE}
# Applying a T-test by providing x and y
x <- sleep$extra[sleep$group == 1]
y <- sleep$extra[sleep$group == 2]
t.test(x, y)

# Applying a T-test with a formula
t.test(extra ~ group, data = sleep)
```

## Example

```{r echo = FALSE}
t.test(extra ~ group, data = sleep)
```

Group 2 shows an increase in sleep length of $\Delta M=1.58$ ($t(17.8)=1.86, p = .08$)

## `wilcox.test()`

-   Calculates a Wilcoxon rank sum test. Also known as Mann Whitney U-Test
-   This test is applied as an alternative to a t-Test when data are assumed to be non-normal distributed.
-   It takes the same arguments as `t.test()`

## Task

-   Take the `sleep` dataset from the previous example.
-   Calculate the `median` and `mad` (Median absolute deviation) for `extra` for each group.
-   Calculate a Wilcoxon test with the `sleep` dataset on the effectiveness of the intervention.

{{< include _stop.qmd >}}

##  {.smaller}

```{r warning=FALSE}
library(dplyr)
sleep %>% group_by(group) %>%
  summarise(
    median = median(extra), 
    mad = mad(extra)
  )
wilcox.test(extra~group, data = sleep)
```

## Task

-   Install the "dslabs" package
-   Take the `gapminder` dataset. This dataset includes health and income outcomes for 184 countries from 1960 to 2016.
-   It includes `infant_mortality`, `region` and `continent`.
-   Calculate the mean and median `infant_mortality` for each continent
-   Calculate a t-test comparing `infant_mortality` between the regions Southern- and Eastern Europe.

{{< include _stop.qmd >}}

## 

```{r eval = FALSE}
install.packages("dslabs")
```

```{r}
library(dslabs)
gapminder %>% group_by(continent) %>% 
  summarise(
    mean = mean(infant_mortality, na.rm = TRUE),
    median = median(infant_mortality, na.rm = TRUE)
  )
```

## 

```{r eval = FALSE}
x <- gapminder %>% filter(region == "Southern Europe") %>% select(infant_mortality)
y <- gapminder %>% filter(region == "Eastern Europe") %>% select(infant_mortality)
t.test(x, y)

# or with pipes
gapminder %>% 
  filter(region %in% c("Southern Europe", "Eastern Europe")) %>%
  t.test(infant_mortality ~ region, data = .)
```

```{r echo = FALSE}
gapminder %>% 
  filter(region %in% c("Southern Europe", "Eastern Europe")) %>%
  t.test(infant_mortality ~ region, data = .)
```
