---
title: "R Basic Concepts"
author: "Jürgen Wilbert"
company: "University of Potsdam"
date: "`r paste0('Version: ', format(Sys.time(), '%d %B %Y'), '<br><br>- arrow keys: move through slides <br>- f: toggle full-screen')`"
output:
  powerpoint_presentation:
    reference_doc: template.pptx
  html_document: default
  pdf_document: default
  beamer_presentation: default
  ioslides_presentation: 
    css: styles.css
    logo: img/ZEIF.png
    widescreen: yes
    df_print: kable
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Goals {.emphasized data-background=img/DEFOCUSED.png data-background-size=cover .flexbox .vcenter}

- Basic concepts of R: library, function, argument 
- Object types: numbers, logic values, characters
- object structures: vectors, data frames
- Sophisticated subsetting
- Import and export data to Microsoft Excel, csv files, SPSS
- Install and update packages (cran, source, and other repositories)
- Organize your work in R Studio (Projects)

## Goals {.emphasized data-background=img/DEFOCUSED.png data-background-size=cover .flexbox .vcenter}

- <div class="red2">**Basic concepts of R: library, function, argument**</div> 
- <div class="red2">**Object types: numbers, logic values, characters**</div>
- <div class="red2">**object structures: vectors, data frames**</div>
- Sophiticated subsetting
- Import and export data to Microsoft Excel, csv files, SPSS
- Install and update packages (cran, source, and other repositories)
- Organize your work in R Studio (Projects)

# Basic concepts

## Functions {.build}

- With a *function* you command the computer to do something.
- *functions* have a function name (e.g., `mean`, `sqrt`).
- *functions* take **arguments** to specify what to do.
- *arguments* have argument names as well.
- functions **always** consist of a function name followed by brackets.

`function_name(argument_name = value, argumentname = value, ...)`

## Examples {.build}

```{r }
# sqrt calculates the square root
sqrt(x = 16)

# You can omit the argument name if it is the first argument
sqrt(16)

# Even without arguments you still need the brackets
date()
```

## Help files

```{r eval=FALSE}
# Function
help("sqrt")

# Short cut
?sqrt
```

... or use the bottom-right help panel in R Studio

## Task 

Take a look at the `mean()` function:  
What *arguments* could be specified?


<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

---

```{r eval=FALSE}
help("mean")
?mean
```

The mean function takes three arguments:

* x : A vector with values
* trim : A fraction (e.g. 0.3) of values that are trimmed from each each end of the vector x 
* na.rm : If TRUE (na.rm = TRUE), missing values (NAs) are removed before calculating the mean

## Operations {.build}

Operations are a special kind of functions that have a shortcut.

```{r results = "hide"}
# function `assign` and the short cut
assign(x = "y", value = 10)

y <- 10
```

```{r results = "hide"}
# function `+`
"+"(e1 = 10, e2 = 10)

10 + 10
```
```{r results = "hide"}
# function `print`
print(x = y)

y
```

## Objects

Objects (also called variables) have an object name and contain data.  
The data are assigned to an object with the `<-` or `=` operator.

```{r}
x <- 10
```

You can see the value(s) of an object with the `print()` function, or by just typing the object name:

```{r results = "hide"}
print(x)
x
```
---

Objects can be used for operators and arguments in functions:

```{r results = "hide"}
x <- 16
y <- 13

x * y
sqrt(x)
```

You can write the return values of a function into a new object:
```{r results = "hide"}
z <- sqrt(x)
z
```
And you can combine these:

```{r results = "hide"}
exp(z) + sqrt(y)

```

## Task {.build}

Assign the values `40 and 24` to the variables `a and b`. Calculate the square root of the sum of `a and b`.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Assign the values `40 and 24` to the variables `a and b`. Calculate the square root of the sum of `a and b`.

```{r}
a <- 40
b <- 24
sqrt(a + b)
```

## Data types {.build}

The data of objects can be numbers, text or TRUE/FALSE values.
These are called **data types**

- **Numeric**: e.g. Integer or decimal numbers `1, 1.35` 
- **Character**: Always between " " or ' ' signs: `"A", 'House'`
- **Logical**:  `TRUE, FALSE`

```{r}
x <- 10
y <- "Hello world!"
z <- FALSE
```

## Data structures {.build}

Data are organized in structures:

>- **Atomic vectors**: A single value
>- **Vectors**: A chain of values
>- **Factors**: Values with assigned labels
>- **List**: A series elements, each one containing one or more (atomic)  vectors
>- **Data Frames**: A list with one vector for each element and all vectors of the same length
>- **Matrix**: A two dimensional table with values of the same data type.
>- **Array**: Like a matrix but with more dimensions.

## Data structures

Data are organized in structures:

- **Atomic vectors**: A single value
- <div class="red2">**Vectors**: A chain of values</div>
- <div class="red2">**Factors**: Values with assigned labels</div>
- **List**: A series elements, each one containing one or more (atomic)  vectors
- <div class="red2">**Data Frames**: A list with one vector for each element and all vectors of the same length</div>
- **Matrix**: A two dimensional table with values of the same data type.
- **Array**: Like a matrix but with more dimensions.

## How to build a vector {.build}

You create a vector with the `c()` function:

```{r}
c(2, 4, 6, 3, 7)
y <- c(2, 4, 6, 3, 7)
y
```

The colon `:` operator creates a numerical sequence:

```{r results = "hide"}
1:10
```

---

You can build a vector of any data type:

```{r}
firstname <- c("Dustin", "Mike", "Will")
curly <- c(TRUE, FALSE, FALSE)
age <- c(9, 11, 10)
```

But do not mix data types in a vector. You will get an error or they are internally changed:

```{r}
age <- c("quite young", 10, 12, "very old")
age
```

## Task {.build}

Create a vector (named `friends` comprising four names of your friends.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Create a vector (named `friends` comprising four names of your friends.

```{r}
friends <- c("Matthias", "Markus", "Thomas", "Christian")
```

## Combining vectors to new vectors {.build}

When an object is a vector it can be reused within the `c()` function to build a new vector:

```{r}
x <- c(3, 5, 7)
c(x, 5, 8, 9)

```

## Combining vectors to new vectors {.build}

Be careful not to confuse an object name with a character:

```{r error = TRUE}
x <- c("A", "B", "C")
c("x", "D", "E", "F")
c(x, "D", "E", "F")
c(A, B, C)

```

## Missing values

A missing value is represented with `NA` (**N**ot **A**vailable).

```{r}
age <- c(9, NA, 11)
name <- c("Tick", "Trick", NA)
age
name

```

## Task {.build}

Create a vector with the values `2, 5, 7, 4, 7, 2, 6`. Calculate the `mean` of these values. (Note: Use the `mean()` function to calculate the mean).

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Create a vector with the values `2, 5, 7, 4, 7, 2, 6`. Calculate the `mean` of these values. (Note: Use the `mean()` function to calculate the mean).


```{r}
x <- c(2, 5, 7, 4, 7, 2, 6)
mean(x)
```

## Task {.build}

Create a vector with the values `2, NA, 7, 4, NA, 2, 6`. Calculate the `mean` of these values.  
(Note: Read through `?mean` if you encounter problems.)

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Create a vector with the values `2, NA, 7, 4, NA, 2, 6`. Calculate the `mean` of these values.  
(Note: Read through `?mean` if you encounter problems.)

```{r}
x <- c(2, NA, 7, 4, NA, 2, 6)
mean(x, na.rm = TRUE)
```


## Selecting elements with square brackets {.build}

```{r}
names <- c("Sheldon", "Leonard", "Penny", "Amy")
names[1]

# Pass a vector to extract multiple elements:
names[c(1,4)]

```

## Task {.build}

Take the vector  
`names <- c("Sheldon", "Leonard", "Penny", "Amy")`  
and reorder it to get the following result:  
`[1] "Sheldon" "Amy"     "Sheldon" "Amy"     "Leonard" "Penny"`

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution 

Take the vector  
`names <- c("Sheldon", "Leonard", "Penny", "Amy")`  
and reorder it to get the following result:  
`[1] "Sheldon" "Amy"     "Sheldon" "Amy"     "Leonard" "Penny"`

```{r}
x <- c(1, 4, 1, 4, 2, 3)
new_order <- names[x]
new_order
```

## A factor {.build .smaller}

A factor is a vector with labels for vector levels. E.g., a vector contains the values 0 and 1 where 0 stands for "with behavioral problems" and 1 stands for ""without behavioral problems".

A factor is created with the `factor()` function.

```{r}
sen <- factor(c(1, 0, 1, 0, 0, 0), levels =  c(0, 1), labels = c("No SEN", "With SEN"))
sen
```
## Task {.build}

Build a factor for `gender` with the labels `male, female, non-binary`. Include a vector for six fictitious gender values.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Build a factor for `gender` with the labels `male, female, non-binary`. Include a vector for six fictitious gender values.

```{r}
gender <- factor(
  c(1, 3, 2, 1, 2, 1), 
  levels = 1:3, 
  labels = c("male", "female", "non-binary")
)
gender
```

## How to build a data frame {.build .smaller}

Data frames are the standard object for storing research data. They contain variables (columns) and cases (rows). A data frame is created with the `data.frame()` function.

```{r}
# For better convenience I have inserted additional linebreaks and spaces
study <- data.frame(
  sen    = c(0, 1, 0, 1, 0, 1),
  gender = c("M", "M", "F", "M", "F", "F"),
  age    = c(12, 13, 11, 10, 11, 14),
  IQ     = c(90, 85, 90, 87, 99, 89)
)
study
```

## Extracting a variable from a data frame

Variables within a data frame are extracted with double square brackets.

```{r}
study[["sen"]]
study[["IQ"]]
```

*Note: an alternative approach is to use the `$` sign: `study$sen`. But we will not use this approach for now.*

## Subsetting a data frame

Specific cases are selected within square brackets: 
`object_name[rows, columns]`.

```{r}
study[5, ]
study[c(2, 6), ]
```

---

```{r}
study[c(2, 6), "IQ"]
study[c(2, 6), c("sen", "IQ")]
```

---

You could also use numbers to address the columns:

```{r}
study[, 2]
study[c(2, 6), c(1, 3)]
```

## Task {.build}

Please create a new data frame (`study2`) comprising the `gender` and `age` variables for the cases 1, 3, and 5 of the `study` data frame.

<div class="red2">**Please stop the video here! Continue after completing the task!**</div> 

## Task - solution

Please create a new data frame (`study2`) comprising the `gender` and `age` variables for the cases 1, 3, and 5 of the `study` data frame.

```{r}
study2 <- study[c(1, 3, 5), c("gender", "age")]
study2
```
