---
title: "dplyr"
subtitle: 'A package tailed towards the manipulation of data frames'
company: "University of Potsdam"
author: "Jürgen Wilbert"
date: "`r paste0('Version: ', format(Sys.time(), '%d %B %Y'), '<br><br>- arrow keys: move through slides <br>- f: toggle full-screen')`"
output:
  ioslides_presentation: 
    css: styles.css
    logo: img/ZEIF.png
    widescreen: yes
    df_print: paged
  html_document: default
  pdf_document: default
  beamer_presentation: default
  powerpoint_presentation:
    reference_doc: template.pptx
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

stop_video <- function() {
  out <- '<div class="red2">**Please stop the video here! Continue after completing the task!**</div>'
  cat(out)
}

```

## dyplyr {.build}

- [`dplyr`](https://dplyr.tidyverse.org/) is a package tailed towards data manipulation.
- `filter()` picks cases based on their values.
- `select()` picks variables based on their names.
- `mutate()` adds new variables that are functions of existing variables
- `summarise()` reduces multiple values down to a single summary.
- `group_by()` allows to perform any operation by grouping variables.
- `arrange()` changes the ordering of the rows.

## `filter()` picks cases based on their values

`filter(.data, ...)` 

```{r eval = FALSE}
# Without %>%:
filter(mtcars, disp > 350 & mpg < 11)

# with %>%
mtcars %>%
  filter(disp > 350 & mpg < 11)
```

Note: You don't have to provide the name of a data object within the dplyr function!

***not***: `filter(mtcars, mtcars$disp > 350 & mtcars$mpg < 11)`

***but***: `filter(mtcars, disp > 350 & mpg < 11)`

## Task

- Take the `starwars` database.
- Filter all cases with `brown` hair color and `female` gender (Hint: variables `hair_color` and `sex`)

```{r results = 'asis', echo = FALSE}
stop_video()
```

***

```{r}
library(tidyverse)
starwars %>% 
  filter(hair_color == "brown" & sex == "female")
```

## `select()` picks variables based on their names

`select(.data, ...)` 

```{r eval = FALSE}
# Either column numbers or column names
mtcars %>% select(1, 3, carb)
# Use the `:` symbol for ranges
mtcars %>% select(1:3, am:carb)
# use `-` to drop variables and keep the others
mtcars %>% select(-am, -(1:3), -(disp:drat))
```

`starts_with()` and `ends_with()` can be used to select variables by parts of their names:

```{r eval = FALSE}
mtcars %>% select(starts_with("d"))
mtcars %>% select(-ends_with("b"))
```

## Task

- Take the `starwars` database.
- Filter for `Human` and select the variables `name, height, and mass` (Hint: variable `species`)

```{r results = 'asis', echo = FALSE}
stop_video()
```

***

```{r}
starwars %>% 
  filter(species == "Human") %>% 
  select(name:mass)
```

## everything() use select to rearrange columns

The `everything()` function helps to rearrange the columns of a data frame 

```{r}
starwars %>% 
  select(name, birth_year, sex, everything())
```


## `mutate()` adds new variables that are functions of existing variables

`mutate(.data, ...)`

```{r eval = TRUE}
mtcars %>%
  mutate(
    efficiency = mpg / disp,
    effect = round(hp / mpg * 100)
  )

```

## Task

- Take the `starwars` database.
- Create a new variable with the body mass index `bmi`. Hint: bmi = kg / m²
- Filter all humans with an bmi >= 26.
- Select `name, bmi, height, and mass`

```{r results = 'asis', echo = FALSE}
stop_video()
```

***

```{r}
starwars %>%
  mutate(bmi = mass/ (height/100)^2) %>%
  filter(bmi >= 26 & species == "Human") %>%
  select(name, bmi, height, mass)
```

## `summarise()` reduces multiple values down to a single summary.

```{r}
mtcars %>%
  summarise(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

## Task {.build}

- Take the `starwars` database.
- Filter `Humans`
- Calculate the mean and the median for height and mass (Note: don't forget to remove NA values: `na.rm = TRUE`)

```{r results = 'asis', echo = FALSE}
stop_video()
```

*** 

```{r}
starwars %>%
  filter(species == "Human") %>%
  summarise(
    mean_height = mean(height, na.rm = TRUE),
    median_height = median(height, na.rm = TRUE),
    mean_mass = mean(mass, na.rm = TRUE),
    median_mass = median(mass, na.rm = TRUE)
  )
```

## `group_by()` allows to perform any operation by grouping variables.

```{r}
mtcars %>% group_by(cyl) %>%
  summarise(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

*** 

```{r}
mtcars %>% group_by(cyl, am) %>%
  summarise(
    mean_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    n = n()
  )
```

## Task {.build}

- Take the `starwars` database.
- Group by sex
- Calculate the median for height and mass
- And the number of cases per group (Hint: `n()`)

```{r results = 'asis', echo = FALSE}
stop_video()
```

***

```{r}
starwars %>%
  group_by(sex) %>%
  summarise(
    median_height = median(height, na.rm = TRUE),
    median_mass = median(mass, na.rm = TRUE),
    n = n()
  )
```

## `arrange()` changes the ordering of the rows

`arrange(.data, ...)`

```{r}
dat <- data.frame(age = c(5,5,6,7,6), sen = c(0, 1, 1, 0, 0), points = c(34, 55, 22, 11, 9))
dat %>%
  arrange(sen, age) 
```

## `desc()` specifies a variable to be arranged descending

```{r}
dat %>%
  arrange(sen, desc(points))
```

## Task

- Take the `starwars` database.
- Calculate the bmi.
- Summarize the median of the bmi grouped by species.
- round the bmi to one decimal.
- Also calculate the `n` for each group.
- Only show cases with `n > 2`.
- Arrange the resulting table. Sort by `n` in descending order.

```{r results = 'asis', echo = FALSE}
stop_video()
```

*** 

```{r}
starwars %>% 
  mutate(bmi = mass / (height / 100)^2) %>% 
  group_by(species) %>% 
  summarise(
    mean_bmi = median(bmi, na.rm = TRUE) %>% round(1), 
    n = n()
  ) %>%
  filter(n >2) %>%
  arrange(desc(n))
           
```
