---
title: "Basic statistics"
subtitle: 'Getting serious with R'
company: "`r if(exists('info')) info$company`"
author: "`r if(exists('info')) info$author`"
date: "`r if(exists('info')) info$date`"
output:
  ioslides_presentation: 
    css: styles.css
    logo: img/ZEIF.png
    widescreen: yes
    df_print: paged
  html_document: default
  pdf_document: default
  beamer_presentation: default
  powerpoint_presentation:
    reference_doc: template.pptx
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

stop_video <- function() {
  #out <- '<div class="red2">**Please stop the video here! Continue after completing the task!**</div>'
   out <- '<div class="red2">**:-)**</div>'
  cat(out)
}

```

## Goals {.emphasized data-background=img/DEFOCUSED.png data-background-size=cover .flexbox .vcenter}

- You know some functions for basic statistical analyses.
- You can create contingency tables.
- You learn to integrate filtering, selecting, grouping, and subsetting of data frames with functions for statistical analyses.

# Descriptive statistics

## Some basic statistical functions

- `min()`, `max()`: Minimum and maximum
- `mean()`, `median()`: Mean and median
- `sd()`, `var()`: Standard deviation and variance
- `mad()`: Median absolute deviation
- `quantile()`: Percentile / quantile

## Task

- Take the `mtcars` dataset
- Create a new variable `lpk` (liters per 100km) from `mpg`\
(Formula: `lpk = 282.5 / mpg`)
- Calculate grouped by cylinders (`cyl`): mean, sd, median, mad, min, max of `lpk`
- Round all values by one decimal
- Hint: Use the `tidyverse` or `dplyr` library

```{r results = 'asis', echo = FALSE}
stop_video()
```

## {.smaller}

```{r}
mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(cyl) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  round(1) 

```

##

What is happening here?

```{r}
mtcars <- mutate(mtcars, 
  transmission = factor(am, labels = c("Manual", "Automatic"))
)
```

##

Now, lets calculate the stats for the transmission type:

```{r}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) 

```

##

... and round the results

```{r eval = FALSE}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  round(1)

```

```{r echo = FALSE}
cat("Error in Math.data.frame(list(transmission = 1:2, mean = c(17.3586131968484, : 
non-numeric-alike variable(s) in data frame: transmission")
```

ups :-( ... what went wrong?

##

Solution:  only `round`  variables that are numeric:

- `is.numeric()` returns a TRUE or FALSE for a vector (`is.numeric(1:5)`; `is.numeric(c("A", "B"))`)  
- `across()` specifies for `mutate()`which variable to select.  

`mutate(across(.cols, .fns, ... ))`:  

- .cols := columns so select  
- .fns := a function to apply  
- ... := arguments to that function  

##

```{r}

mtcars %>% 
  mutate(lpk = 282.5 / mpg) %>% 
  group_by(transmission) %>% 
  summarise(
    mean = mean(lpk),
    sd = sd(lpk),
    median = median(lpk),
    mad = mad(lpk),
    min = min(lpk),
    max = max(lpk)
  ) %>%
  mutate(across(is.numeric, round, 1))

```

# Tables

## Tables in base R 

- `table()`: Shows frequencies of nominal scaled variables.
- `prop.table()`: Calculates proportions from frequency tables.
- `addmargins()`: Adds margins to tables.

## Example one dimensional {.smaller}

```{r}
tab <- table(mtcars$cyl) # frequencies
tab
prop.table(tab) # proportions
prop.table(tab) * 100 # percentages 
```

## Example two dimensional {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
tab
prop.table(tab) * 100
```

## Example two dimensional with percentages by rows and columns {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
prop.table(tab, margin = 1) * 100  # sum of each row is 100%
prop.table(tab, margin = 2) * 100 # sum of each column is 100%
```

## Example with added margins {.smaller}

```{r}
tab <- table(mtcars$cyl, mtcars$am)
tab <- prop.table(tab, margin = 2) * 100
addmargins(tab)
```

## Task

- Take the `heights` dataset from the `dslabls` library.  
- Create a cross table depicting: The percentage of females and males that are above and below a height of 170 cm.  

##

```{r}
library(dslabs)

heights$category[heights$height * 2.54 <= 170] <- "% <= 170 cm"
heights$category[heights$height * 2.54 > 170] <- "% > 170 cm"

tab <- table(heights$sex, heights$category)
tab <- prop.table(tab, margin = 1) * 100
tab <- round(tab, 1)
tab
```


## Tables with dplyr

```{r}
mtcars %>% 
  group_by(cyl, am) %>% 
  summarise(n = n())

```

## Tables with dplyr: `pivot_wider()` {.smaller}

`pivot_wider()`:  
creates separate variables from the levels of a factor variable and the values of a second variable


## {.smaller} 

```{r}
mtcars %>% group_by(cyl, am) %>% summarise(n = n())
```


```{r}
mtcars %>% group_by(cyl, am) %>% summarise(n = n()) %>%
  pivot_wider(names_from = "am", values_from = "n")

```


## Tables with dplyr: A bit nicer!

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(n = n()) %>%
  pivot_wider(names_from = "am", values_from = "n") %>%
  rename("Cylinders" = "cyl")

```

## Task

Recreate the previous task. This time using dplyr:

- Take the `heights` dataset from the `dslabls` library.  
- Create a cross table depicting: The percentage of females and males that are above and below a height of 170 cm.  

This is a hard task ...

Hint 1: `ifelse(height * 2.54 <= 170, "size_a", "size_b")`  
Hint 2:  
`size_a / (size_a + size_b) * 100`   
`size_b / (size_a + size_b) * 100`  
are the row proportions.

##

```{r}
library(dslabs)

heights %>%
  mutate(category = ifelse(height * 2.54 <= 170, "size_a", "size_b")) %>% 
  group_by(sex, category) %>%
  summarise(n = n()) %>% 
  pivot_wider(names_from = "category", values_from = "n") %>%
  mutate(
    "% <= 170" = size_a / (size_a + size_b) * 100,
    "% > 170" = size_b / (size_a + size_b) * 100,
    across(is.numeric, round, 1)
  ) %>%
  select(1, 4, 5)

```

## Tables with dplyr: For values of a third variable

```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(
    n = n(), 
    M = mean(mpg), 
    SD = sd(mpg)
  ) 
```

## 

`pivot_wider()` can take values from several variables:


```{r}
mtcars %>% 
  mutate(am = factor(am, labels = c("Manual", "Automatic"))) %>%
  group_by(cyl, am) %>% 
  summarise(
    n = n(), M = mean(mpg), SD = sd(mpg)
  ) %>%
  pivot_wider(names_from = "am", values_from = c("n", "M", "SD")) %>%
  round(1)

```

## Task

- Take the `storms` dataset and calculate n and mean of the wind speed (`wind`) by month (`month`) and storm classification (`status`) in a crosstable.

##

```{r}
storms %>% 
  group_by(month, status) %>%
  summarise(
    n = n(),
    M = mean(wind, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = "status", values_from = c("n", "M")) %>%
  round()
  
```



# Inferential statistics

## Some statistical functions:

- `t.test()`: Calculating a t-test.
- `wilcox.test()`: Calculating Wilcox test / U-Test.
- `chisq.test()`: Calculating a Pearson $X^2$-test.

- Some more we will not address today:
    - `lm()`: Regression analyses
    - `cor.test()`: Calculating a correlation test.
    - `binom.test()`: Binomial test.
    - `fisher.test()`: Fisher exact test for count data.
    - `ks.test()`: One and two sample Kolmogorov-Smirnov Tests.
    - `shapiro.test()`: Shapiro-Wilk Normality Test.
    - `aov()`: Analysis of variance

## `chisq.test()`

- *Pearson's* $X^2$ test is a very versatile test to calculate whether a distribution of observed frequencies equates expected frequencies.  
- A very common application is to test whether frequencies of observations in two variables are related.  (E.g. number of patients that improved in a treatment vs. control group).
- The `chisq.test()` functions takes a two dimensional frequency table and calculates a $X^2$ test.

## Example with data from an intervention study {.smaller}

```{r}
# get an external data example
dat <- read.csv("https://goo.gl/j6lRXD")
# Create a two distribution table
tab <- table(dat$treatment, dat$improvement)
tab
# Test for non random distribution:
chisq.test(tab) # Alternatively: chisq.test(dat$treatment, dat$improvement)
```

Significantly more patients improved in the treatment condition ($X^2(1)=4.7, p < .05$)

## Task

- Take the `starwars` dataset from the `tidyverse` package.
- Use `table()` to get the distribution `eye_color` by `hair_color`
- Apply the `chisq.test()` to calculate a X² test to test for an even distribution.

```{r results = 'asis', echo = FALSE}
stop_video()
```

##

```{r warning=FALSE}
tab <- table(starwars$eye_color, starwars$hair_color)
chisq.test(tab)
```

## `t.test()`

- *Student's t-test* analysis whether two samples originate from the same normal distribution.  
- It is used to test for mean differences in two groups.  
- ***Arguments*** of the `t.test()` function:
    - `x` and `y`: Each variable provides data from a samples.
    - `formula`: If you have one vector with all data (e.g. `values`) and a second vector with grouping information (e.g. `group`) use: `values ~ group`. 
    - `data`: If you work with `formula`: The name of a data frame.
    - `paired`: If set `TRUE`, the test assumes repeated measures of one sample instead of two independent samples.



## Example

Does one of two drugs increase hours of sleep better?

`extra`: change in sleep duration, `group` : drug given

```{r}
sleep
```


## Example

```{r eval = FALSE}
# Applying a T-test by providing x and y
x <- sleep$extra[sleep$group == 1]
y <- sleep$extra[sleep$group == 2]
t.test(x, y)

# Applying a T-test with a formula
t.test(extra ~ group, data = sleep)
```

## Example

```{r echo = FALSE}
t.test(extra ~ group, data = sleep)
```

Group 2 shows an increase in sleep length of $\Delta M=1.58$ ($t(17.8)=1.86, p = .08$)

## `wilcox.test()`

- Calculates a Wilcoxon rank sum test. Also known as Mann Whitney U-Test
- This test is applied as an alternative to a t-Test when data are assumed to be non-normal distributed.
- It takes the same arguments as `t.test()`

## Task

- Take the `sleep` dataset from the previous example.
- Calculate the `median` and `mad` (Median absolute deviation) for `extra` for each group.
- Calculate a Wilcoxon test with the `sleep` dataset on the effectiveness of the intervention.

```{r results = 'asis', echo = FALSE}
stop_video()
```

## {.smaller}

```{r warning=FALSE}
library(dplyr)
sleep %>% group_by(group) %>%
  summarise(
    median = median(extra), 
    mad = mad(extra)
  )
wilcox.test(extra~group, data = sleep)
```

## Task

- Install the "dslabs" package
- Take the `gapminder` dataset. This dataset includes health and income outcomes for 184 countries from 1960 to 2016.
- It includes `infant_mortality`, `region` and `continent`.
- Calculate the mean and median `infant_mortality` for each continent
- Calculate a t-test comparing `infant_mortality` between the regions Southern- and Eastern Europe.

```{r results = 'asis', echo = FALSE}
stop_video()
```

## 

```{r eval = FALSE}
install.packages("dslabs")
```

```{r}
library(dslabs)
gapminder %>% group_by(continent) %>% 
  summarise(
    mean = mean(infant_mortality, na.rm = TRUE),
    median = median(infant_mortality, na.rm = TRUE)
  )
```

## 

```{r eval = FALSE}
x <- gapminder %>% filter(region == "Southern Europe") %>% select(infant_mortality)
y <- gapminder %>% filter(region == "Eastern Europe") %>% select(infant_mortality)
t.test(x, y)

# or with pipes
gapminder %>% 
  filter(region %in% c("Southern Europe", "Eastern Europe")) %>%
  t.test(infant_mortality ~ region, data = .)
```

```{r echo = FALSE}
gapminder %>% 
  filter(region %in% c("Southern Europe", "Eastern Europe")) %>%
  t.test(infant_mortality ~ region, data = .)
```


