---
title: "Control functions"
company: "`r if(exists('info')) info$company`"
author: "`r if(exists('info')) info$author`"
date: "`r if(exists('info')) info$date`"
output:
  ioslides_presentation: 
    css: styles.css
    logo: img/ZEIF.png
    widescreen: yes
  html_document: default
  pdf_document: default
  beamer_presentation: default
  powerpoint_presentation:
    reference_doc: template.pptx
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

stop_video <- function() {
  #out <- '<div class="red2">**Please stop the video here! Continue after completing the task!**</div>'
  out <- '<div class="red2">**:-)**</div>'
  cat(out)
}

```

## if control structures

if-else control structures are very helpful for solving complex conditional operations:

`if (condition) expression`

example:

`if (age >= 18) adult <- TRUE`

The expression can take multiple lines which are enclosed with braces:

`if (age >= 18 && gender == "male") {`  
`  adult <- TRUE`  
`  n_males <- n_males + 1`  
`}`

## if-else control structures

Sometimes, the `if` structure is extended with `else`:

`if (condition) expression1 else expression2`

example:

`if (age >= 18) out <- "full-aged" else out <- "underage"`

This can also take multiple commands:

`if (age >= 18 && gender == "male") {`  
`  adult <- TRUE`  
`  n_males <- n_males + 1`  
`} else if (age >= 18 && gender == "female") {` 
`  adult <- TRUE`  
`  n_females <- n_females + 1`  
`} else if (age < 18) {`    
`  adult <- FALSE`  
`}`  

## 

The `if` structure is not vectorized.  
That is, the condition must return a single logical TRUE or FALSE value:

```{r warning=TRUE}
age <- c(18, 15)
if (age >= 18) cat("Adult!") else cat("Underage!")
```

## ifelse

`ifelse` is a vectorized version of `if()` `else` which returns a vector:

`ifelse(condition, expression_true, expression_false)`

Example:

```{r}
age <- c(15, 18, 14, 12)
ifelse(age >= 18, "adult", "underage")
```

## Task: Read the code

What is the object `results` after executing the code?

```{r eval = FALSE}
age <- c(15, 19, 21, 18, 17, 23)
cut_full_aged <- 18
language <- "german"
if (language == "english") {
  label_adult <- "adult"
  label_underage <- "underage"
} else if (language == "german") {
  label_adult <- "Vollj채hrig"
  label_underage <- "Minderj채hrig"
} else {
  stop("Unknown language!")
}

results <- ifelse(age < cut_full_aged, label_underage, label_adult)

```

##

```{r}
age <- c(15, 19, 21, 18, 17, 23)
cut_full_aged <- 18
language <- "german"
if (language == "english") {
  label_adult <- "adult"
  label_underage <- "underage"
} else if (language == "german") {
  label_adult <- "Vollj채hrig"
  label_underage <- "Minderj채hrig"
} else {
  stop("Unknown language!")
}

results <- ifelse(age < cut_full_aged, label_underage, label_adult)

results
```

## Task: Read the code

What is the object `results` after executing the code?

```{r eval = FALSE}
bj <- 1900 + 1:5 * 20
results <- ifelse(bj > 1996, "Generation Z", 
                  ifelse(bj > 1980, "Millenial", 
                         ifelse(bj > 1964, "Generation X", 
                                ifelse(bj >1945, "Babyboomer", 
                                       ifelse(bj > 1927, "Silent generation", "Unknown")))))
                       

```

##

```{r}
bj <- 1900 + 1:5 * 20
results <- ifelse(bj > 1996, "Generation Z", 
                  ifelse(bj > 1980, "Millenial", 
                         ifelse(bj > 1964, "Generation X", 
                                ifelse(bj >1945, "Babyboomer", 
                                       ifelse(bj > 1927, "Silent generation", "Unknown")))))
                       
results
```

## For loops

With a `for` loop you can iterate the values of a vector:

**for** (*variable* **in** *vector*) *expression*

example:

```{r}
for (i in 1:10) print(i)
```

## 

An expression can have multiple lines enclosed in braces:

```{r}
adult <- 0
minor <- 0
for(i in c(20, 12, 15, 16, 18)) {
  if (i >= 18) adult = adult + 1
  if (i < 18) minor = minor + 1
}
cat("Adults:", adult)
cat("Minors:", minor)
```

## 

Oftentimes `for` loops are used to iterate multiple vectors.

Can you read this code and predict the object `comparison` ? 

```{r eval = FALSE}
age1 <- c(20, 12, 15, 16, 18)
age2 <- c(17, 13, 17, 16, 20)
comparison <- c()
for(i in 1:length(age1)) {
  if (age1[i] < age2[i]) tmp <- "younger"
  if (age1[i] > age2[i]) tmp <- "older"
  if (age1[i] == age2[i]) tmp <- "equal"
  comparison <- c(comparison, tmp)
}

comparison

```

## 

You can use `for` loops to iterate through a dataframe:

```{r eval = FALSE}

mtcars$impact <- "average"

for(i in 1:nrow(mtcars)) {
  
  if (mtcars$mpg[i] < 18 && mtcars$cyl[i] == 8 && mtcars$wt[i] > 4) 
    mtcars$impact[i] <- "Big waster!"
  if (mtcars$mpg[i] > 22 && mtcars$cyl[i] == 4 && mtcars$wt[i] < 2.59) 
    mtcars$impact[i] <- "Light eco!"
}

mtcars[, c("mpg", "cyl", "wt", "impact")]

```

## 

```{r echo = FALSE}
mtcars$impact <- "average"

for(i in 1:nrow(mtcars)) {
  
  if (mtcars$mpg[i] < 18 && mtcars$cyl[i] == 8 && mtcars$wt[i] > 4) 
    mtcars$impact[i] <- "Big waster!"
  if (mtcars$mpg[i] > 22 && mtcars$cyl[i] == 4 && mtcars$wt[i] < 2.59) 
    mtcars$impact[i] <- "Light eco!"
}

mtcars[, c("mpg", "cyl", "wt", "impact")]
```


## Functions

Repeated operations can be put into functions.

*functionname* <- **function**(*argument*, *argument*, *argument*, ...) {*expression*}

Example:

```{r}
w20_die <- function(dice) {
  if (dice < 1 || dice > 20) return("This is not a decent W20 die!")
  if (dice == 20) return("Critical hit!!")
  if (dice == 1) return("Epic fail")
  dice
}
```

`return()` stops the function and returns an object. If no return is defined, the function will return the last object that is printed.

##

```{r}
w20_die <- function(dice) {
  if (dice < 1 || dice > 20) return("This is not a decent W20 die!")
  if (dice == 20) return("Critical hit!!")
  if (dice == 1) return("Epic fail")
  dice
}
w20_die(20)
w20_die(3)
w20_die(1)
w20_die(0)
```

## Task Read the code

What happens here?

```{r}
percentage <- function(x, na.rm = TRUE) {
  if (!is.logical(x)) stop("Please provide a logical vector.")
  if (na.rm) x <- x[!is.na(x)]
  sum(x) / length(x) * 100
}

var <- c(12, 16, 19, 20, 11)

percentage(var >= 18)

```

## Task {.smaller}

```{r eval = FALSE}
w20_die <- function(dice) {
  if (dice < 1 || dice > 20) return("This is not a decent W20 die!")
  if (dice == 20) return("Critical hit!!")
  if (dice == 1) return("Epic fail")
  dice
}
```

- Vectorize the `w20_die` function so it will return a vector with results when you provide a vector with numbers as an argument:


```{r echo = FALSE}
w20_die <- function(dice) {
  out <- dice
  for(i in 1:length(dice)) {
    if (dice[i] < 1 || dice[i] > 20) out[i] <- "This is not a decent W20 die!"
    if (dice[i] == 20) out[i] <- "Critical hit!!"
    if (dice[i] == 1) out[i] <- "Epic fail"
  }
  out
}
```

```{r}
results <- c(2, 16, 20, 1, 0)
w20_die(results)
```


##

```{r eval = FALSE}
w20_die <- function(dice) {
  out <- c()
  for(i in 1:length(dice)) {
    out[i] <- dice[i]
    if (dice[i] < 1 || dice[i] > 20) out[i] <- "This is not a decent W20 die!"
    if (dice[i] == 20) out[i] <- "Critical hit!!"
    if (dice[i] == 1) out[i] <- "Epic fail"
  }
  out
}
```

