---
title: "Create new variables in data frames"
date: "`r info$date`"
author: "`r info$author`"
format: 
  revealjs:
    logo: /img/ZEIF.png
    scrollable: true
    theme: [style.scss]
    footer: "Jürgen Wilbert - Introduction to R"
title-slide-attributes:
    data-background-color: "#a9bfd1"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Create a new variable in a dataset with missing values

Please copy this syntax into R-Studio and execute:

```{r}
study <- data.frame(
  sen    = factor(c(0, 1, 0, 1, 0, 1), labels = c("no sen", "sen"), levels = 0:1),
  gender = factor(c(1,2,2,1,2,1), labels = c("Male", "Female"), levels = 1:2),
  age  = c(150, 156, 138, 126, 136, 162),
  IQ     = c(90, 85, 90, 87, 99, 89),
  Q1     = c(1, 5, 3, 4, 2, 3),
  Q2     = c(2, 5, 3, 4, 5, 2),
  Q3     = c(1, 5, 3, 5, 2, 1)
)
```

## Create a new variable in a dataset {.smaller}

With missing values:

```{r}
study$Q4 <- NA
```

With specific values:

```{r}
study$Q5 <- c(5, 1, 4, 2, 1, 3)
study
```

## Create new variables from existing ones

```{r}
study$age2 <- study$age / 12
study
```

## Drop a variable

```{r}
study <- subset(study, select = -age2)
# or: study[, -10]
study
```

## Task

Create a new variable `Q_sum_1_3` as the sum of `Q1` to `Q3`.

{{< include _stop.qmd >}}

---

```{r}
study$Q_sum_1_3 <- with(study, Q1 + Q2 + Q3)

# or:

study$Q_sum_1_3 <- study$Q1 + study$Q2 + study$Q3
study
```

## Task

Create the variables `age_year` and `age_month` where both variables do not have decimals 

(tip: `trunc` function or the `%%` modulo operator and the integer division `%/%` operator. Use the help function if needed)

{{< include _stop.qmd >}}

---

```{r}
study$age_year <- trunc(study$age /12)
study$age_month <- study$age - (study$age_year * 12)

# or

study$age_year <- study$age %/% 12
study$age_month <- study$age %% 12

study
```

## The apply function {.smaller}

The `apply` function applies a function to every row or column of a data frame.

1. The first argument is the data frame object.

2. The second argument is the margin (1 for rows and 2 for columns).

3. The third argument is a function name (e.g. mean).

4. Further arguments are arguments to the function provided in 3.

```{r}
# An example:
study$Q_sum_1_2 <- apply(study[, c("Q1", "Q2")], 1, sum)
study
```

## Task

Create a new variable `Q_sum` with the sum of Q1 to Q5 (sums should be build when NAs are in the data)

---

```{r}
study$Q_sum <- apply(subset(study, select = Q1:Q5), 1, sum, na.rm = TRUE)

```


