---
title: "Data analyses using R - schedule"
company: "`r if(exists('info')) info$company`"
author: "`r if(exists('info')) info$author`"
date: "`r if(exists('info')) info$date`"
output:
  ioslides_presentation:
    logo: img/ZEIF.png
    smaller: yes
  powerpoint_presentation:
    reference_doc: template.pptx
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
schedule <- readxl::read_xlsx("schedule.xlsx")
```

# Schedule

---

```{r}
knitr::kable(schedule)
```

## Schedule 2

| Date | Topic |
|------|-------|
22.10. | Installation R and R Studio |
|05.11. | Basics of R, part 1 |
|12.11. | Basics of R, part 2 |
|19.11. | Markdown and Rmarkdown |
|26.11. | Data cleaning and manipulation with tidyverse |
|03.12. | Data visualization with ggplot |
|10.12. | Regression analyses with R |
|07.01. | Scale construction and Item analyses in R (part 1) |
|14.01. | Scale construction and Factor analyses in R (part 2)  |
|21.01. | Going retro: SPSS introduction, part 1 |
|28.01. | Going retro: SPSS introduction, part 2 |
